\setdictum{%
  B-splines are not enough!%
}{%
  In a talk at the 2017 SIAM Conference on\\
  Computational Science and Engineering%
}
% SIAM CSE 2017, Minisymposium MS154 "Flooding the Cores--Computing Flooding
% Events on Modern Architecture--Part I of II",
% Craig Michoski, UT Austin,
% "Scaling at Exascale in Blended Isogeometric, Discontinuous
% Galerkin, and Particle-in-Cell Approaches"

\chapter{Hierarchical B-Splines and Sparse Grids}
\label{sec:20sparseGrids}

\todo{Write short intro to SGs? Overlap with Chap. 1?}

The aim of this chapter is to define hierarchical B-splines on sparse grids.
It is structured as follows:
\Cref{sec:21sparseGrids} provides a consistent notational framework
for the definition of sparse grids with general basis functions.
The reason not to employ specific bases such as the common hat functions
or B-splines of higher degrees is two-fold:
First, we will define various new ``flavors'' of B-splines,
which is easier if the basis is left open.
Second, most of the statements and theorems that we will make in this
thesis will hold for general basis functions
(in some cases with additional assumptions)
and not just for B-splines.
Next, \cref{sec:22bSplines} defines hierarchical B-splines for both
uniform and non-uniform knot sequences.
Finally, we see in \cref{sec:23boundary} that the boundary behavior
of the classical uniform B-spline basis is problematic,
which is mitigated by incorporating not-a-knot boundary conditions
into the B-spline basis.

\input{21sparseGrids}
\input{22bSplines}
\input{23boundary}
