\setdictum{%
  B-splines are not enough!%
}{%
  In a talk at the 2017 SIAM Conference on\\
  Computational Science and Engineering%
}
% SIAM CSE 2017, Minisymposium MS154 "Flooding the Cores--Computing Flooding
% Events on Modern Architecture--Part I of II",
% Craig Michoski, UT Austin,
% "Scaling at Exascale in Blended Isogeometric, Discontinuous
% Galerkin, and Particle-in-Cell Approaches"

\chapter{Hierarchical B-Splines}
\label{chap:30BSplines}



\lettrine{I}{n} the last chapter,
we repeated the definition of sparse grids for
arbitrary tensor product basis functions.
The piecewise linear ``hat function'' basis $\bspl{l,i}{1}{}$
served as the motivation to omit hierarchical subspaces with
little contribution to the overall approximation quality.
However, the hat function basis is not continuously differentiable.
This has two implications.

\todo{mention order of splines on SGs?}

The first implication is that the approximation order of hat functions
is lower than the order of other basis function types
such as higher-degree splines \cite{Sickel11Spline}
or the piecewise polynomial basis by Bungartz \cite{Bungartz98Finite}.
The second implication, which also applies to Bungartz's basis,
is that we cannot compute globally continuous gradients of the
interpolant of a smooth objective function,
if we use non-smooth basis functions.
However, the availability of gradients is
essential in the application of gradient-based optimization,
which we target in this thesis.
In this chapter, we therefore define a hierarchical and
higher-order B-spline basis
(generalizing the well-known hat functions)
to obtain both higher-order approximations
as well as continuous gradients and Hessians.

The first to study B-splines was Isaac Schoenberg in 1946
\cite{Schoenberg46Contributions},
but he claimed that they had already been known to Laplace
\cite{Boor76Splines}.
Research and industry recognized the possibilities of B-splines when
the \fem, which is still the most significant application of B-splines,
emerged in the 1960s.
Carl de~Boor pioneered B-splines, developed basic algorithms, and
proved fundamental theoretical results \cite{Boor72Calculating}.
Researchers have applied B-splines in a diverse set of fields such as
\fem \cite{Hoellig03Finite} and \iga \cite{Hoellig12Finite},
geometric modeling with \nurbs
\multicite{Cohen01Geometric,Hoellig13Approximation},
financial mathematics \cite{Pflueger10Spatially},
molecular and atomic physics
\multicite{Bachau01Applications,McCurdy04Implementation},
and numerous other scientific and industrial areas.

In this chapter, we define hierarchical B-splines on sparse grids.
The chapter is divided into two sections:
First, we define hierarchical B-splines for both
uniform and non-uniform knot sequences in \cref{sec:31standardBSplines}.
Second, we learn in \cref{sec:32notAKnot} that the boundary behavior
of the classical uniform B-spline basis is problematic.
Incorporating not-a-knot boundary conditions into the B-spline basis
mitigates the problems caused by the boundary behavior.

\Cref{sec:311uniform} is a repetition of the definition
of nodal B-splines \multicite{Hoellig03Finite,Hoellig13Approximation} and
hierarchical B-splines \multicite{Pflueger10Spatially,Valentin14Hierarchische}.
Original contributions of the thesis in this chapter are the proof of
the linear independence of hierarchical B-splines in
\cref{sec:312proofHierarchicalSplitting}
(improved version of \cite{Valentin14Hierarchische},
published in \cite{Valentin16Hierarchical}),
the modified hierarchical Clenshaw-Curtis B-splines in
\cref{sec:314nonUniform} and
the hierarchical not-a-knot B-spline basis in \cref{sec:32notAKnot}.
\todo{add citation if Vazipfl has been published}



\input{31standardBSplines}
\input{32notAKnot}

\cleardoublepage
