\addchap{Preface}

Before I start, I want to thank my advisor Prof.\ Dr.\ Dirk Pflüger.
It is his ideas and his valuable input that have driven me in
my time as a PhD student.
I also thank the other members of the
examination board for their willingness to examine my thesis,
namely \theexamineri{}, \theexaminerii{}, and \theexamineriii{}.

Similarly, I thank
Peter Schober (Goethe University Frankfurt),
Daniel Hübner (FAU Erlangen-Nürnberg), and
Dr.\ Michael Sprenger (SimTech/University of Stuttgart) for
collaborating and the enlightening discussions.
I am also grateful for the exciting time with the whole group of
SSE (Simulation Software Engineering) and
SGS (Simulation of Large Systems),
for which I want to thank all past and current PhD students and postdocs of
Dirk Pflüger and Prof.\ Dr.\ Miriam Mehl.
I thank
Carolin Schober,
Henriette Röger, and
Ralf Diestelkämper
for throughly proofreading
parts of drafts of this thesis.

Probably the most important role for the success of my PhD thesis
has played my family.
Without their mental support and distraction from the daily work,
I doubt that this thesis would have been possible.

Likewise, I am very grateful for the financial support from
the \foreignlanguage{ngerman}{Juniorprofessurenprogramm} of the
\foreignlanguage{ngerman}{Baden-Württemberg Stiftung}.
I thank the SimTech Cluster of Excellence for supporting
my three-month research stay in Canberra, Australia.

In addition, I want to thank the open-source community for making it possible to
write this thesis in an aesthetically sophisticated manner.
The list of software that was used to write this thesis includes
\LaTeX, Lua\LaTeX, Bib\LaTeX,
\scalebox{0.9}{\KOMAScript}, Ti\emph{k}Z, Python, Matplotlib,
and many more.

\label{page:preface}
Now, I wish that you, dear reader, obtain as much insight as possible
while reading the remaining
\pagedifference{page:preface}{LastPage} pages of this thesis.

Enjoy!

\vspace{1em}

\noindent
Stuttgart, \thedate

\noindent
\theauthor

\cleardoublepage
